---
title: "Analysis of Shootings and Firearm Discharges in Toronto"
subtitle: "A Comprehensive Examination Reveals Significant Time and Locational Impacts on Incident Severity"
author:
  - name: Xinze Wu
    affiliation: University of Toronto
    email: kerwin.wu@mail.utoronto.ca
thanks: "Code and data are available at: https://github.com/ke3w/Data-Analysis-of-Shootings-and-Firearm-Discharges-in-Toronto"
date: today
date-format: long
abstract: "This paper analyzes the impact of time and location on the severity of shootings and firearm discharges in Toronto, utilizing a Gradient Boosting Machine model. The findings highlight significant variations across different times and locations, emphasizing the need for targeted policy interventions. This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives.The narrative structure of this analysis follows the guidelines provided in 'Telling Stories with Data' [@tellingstories]."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gbm)
library(caret)
library(ggplot2)

#### Load saved model ####
model <- readRDS("../models/first_model.rds")

# Load the cleaned and preprocessed data for further analysis or prediction
analysis_data <- read_csv("../data/02-analysis_data/analysis_data.csv")

analysis_data <- analysis_data %>%
  mutate(weighted_score = death * 2 + injuries)  # Adjust the greater importance of 'death' over 'injuries'

```

# Introduction

The safety and security of urban areas are of paramount importance to their residents and governance structures. In Toronto, the occurrence of shootings and firearm discharges poses significant challenges to public safety. Understanding the patterns and determinants of these violent events is crucial for developing effective interventions and policies. This paper employs a Gradient Boosting Machine model to analyze how time-related factors and geographic locations influence the severity of shooting incidents in Toronto.

## Estimand

This study aims to estimate the effect of time (day of week, hour, and time range) and location (neighbourhood and police division) on the severity of shootings and firearm discharges in Toronto, measured by a weighted score combining deaths and injuries.

## Importance of the Study

The results of this study are crucial for informing public safety strategies and police resource allocation in Toronto, aiming to reduce the incidence and severity of violent firearm-related incidents.

# Data {#sec-data}

Data for this study is sourced from Open Data Toronto[@OpenDataToronto]: <https://open.toronto.ca/dataset/shootings-firearm-discharges/>, detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.

## Data Cleaning

Data cleaning was performed using R[@citeR] packages such as `tidyverse`[@tidyverse] and `lubridate`. The `occ_date` variable was converted to a numeric format to fit the model requirements, and categorical variables were transformed using appropriate factor conversions.

## Variables

### Outcome Variable

-   `Weighted Score`: Combines the number of deaths and injuries, weighting deaths twice as heavily(weighted_score = death \* 2 + injuries).

### Predictor Variables

-   **Time Factors**:

-   `occ_date`(date): Date of Offence Occurred

-   `occ_dow` (day of the week): Day of the Month Offence Occurred

-   `occ_time_range` (morning, afternoon, evening):Time Range of Day Offence Occurred

-   **Location Factors**:

-   `neighbourhood_158:`Name of Neighborhood using City of Toronto's new 158 neighborhood structure

-   `division`: Police division where offence occurred

## Data visualization

### Weighted Score Over Time

```{r}
#| warning: false
#| message: false
#| label: fig-date
#| fig_caption: true
#| fig-cap: "line graph shows the relationship between weighted score over date"
#| echo: false

ggplot(analysis_data, aes(x = occ_date, y = weighted_score)) +
  geom_line(group = 1, color = "blue") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Weighted Score Over Time", x = "Date", y = "Weighted Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

This plot @fig-date illustrates the variation in the weighted scores of shootings and firearm discharges in Toronto from 2004 to 2025. The weighted score, which combines the number of deaths and injuries with deaths being given a higher weight, is plotted against the date of incidents. Notably, the plot reveals several significant spikes, indicating periods with higher shooting severity. These peaks might correlate with specific events or changes in local circumstances, warranting further investigation. The majority of the time, however, the weighted scores remain relatively low, suggesting sporadic rather than consistent patterns of high-severity incidents. This visualization highlights the dynamic nature of crime severity over time and underscores the importance of continuous monitoring and analysis to understand the underlying trends and triggers.

### Weighted Score by Day of week

```{r}
#| warning: false
#| message: false
#| label: fig-dow
#| fig_caption: true
#| fig-cap: "Bar chart of the weighted score by the day of week"
#| echo: false

ggplot(analysis_data, aes(x = occ_dow, y = weighted_score)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Weighted Score by Day of Week", x = "Day of Week", y = "Weighted Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The bar chart @fig-dow illustrates the weighted scores of shootings and firearm discharges in Toronto, broken down by day of the week. Notably, the weekend days (Saturday and Sunday) along with Monday, show elevated scores, suggesting a higher incidence of severe incidents during these days. This trend could be attributed to increased social activities and gatherings during the weekend, potentially leading to more conflicts or accidents. Conversely, the mid-week days (Tuesday and Wednesday) exhibit notably lower scores, indicating fewer severe incidents.

### Weighted Score by Day of Year

```{r}
#| warning: false
#| message: false
#| label: fig-doy
#| fig_caption: true
#| fig-cap: "line graph of the weighted score by the day of year"
#| echo: false

ggplot(analysis_data, aes(x = occ_doy, y = weighted_score)) +
  geom_smooth(method = "loess", colour = "darkgreen") +
  labs(title = "Weighted Score by Day of Year", x = "Day of Year", y = "Weighted Score")

```

The line graph @fig-doy visualizes the trend of shootings and firearm discharges in Toronto across different days of the year. The plot shows a clear seasonal pattern, with peaks generally occurring around mid-year and towards the end, particularly noticeable in the increase during summer and winter months. This could correlate with seasonal activities and social behaviors such as holidays and outdoor gatherings that might contribute to increased incidents. The valleys observed during early spring and late fall might reflect quieter periods with fewer such gatherings.

### Weighted Score by Hour of Day

```{r}
#| warning: false
#| message: false
#| label: fig-hour
#| fig_caption: true
#| fig-cap: "Bar chart shows the weighted score by the hour of day"
#| echo: false

ggplot(analysis_data, aes(x = occ_hour, y = weighted_score)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(title = "Weighted Score by Hour of Day", x = "Hour", y = "Weighted Score")


```

The bar graph @fig-hour presents the distribution of shootings and firearm discharges in Toronto by hour. This visualization demonstrates a notable peak in incidents during the very early hours of the day, specifically around midnight, with a dramatic decline shortly after. A secondary, but less intense, peak occurs in the evening hours, spanning from 8 PM to midnight.

This pattern suggests that incidents tend to happen more frequently during late-night hours, possibly linked to social and recreational activities during these times, or reduced visibility and police presence making these hours more conducive to criminal behavior. Conversely, the hours from early morning to mid-afternoon show markedly lower incidents, reflecting perhaps quieter public activity and higher visibility and vigilance during daylight hours.

### Weighted Score by Time Range

```{r}
#| warning: false
#| message: false
#| label: fig-range
#| fig_caption: true
#| fig-cap: "Bar chart shows the weighted score by the time range"
#| echo: false

ggplot(analysis_data, aes(x = occ_time_range, y = weighted_score, fill = occ_time_range)) +
  geom_bar(stat = "identity") +
  labs(title = "Weighted Score by Time Range", x = "Time Range", y = "Weighted Score")

```

This bar graph @fig-range illustrates the severity of shootings and firearm discharges in Toronto across different time range of the day. The time ranges are categorized into morning, afternoon, evening, and night.

The graph reveals a stark increase in incident severity during the night, which towers over the scores recorded for other times of the day. This is followed by the evening, which also shows a considerably high level of incident severity. The overall trend of this graph is highly corresponding to the bar chart of weighted score over hour of the day, basically, this graph provides a more straightforward relationship between the weight score over different periods in a day.

### Weighted Score by Neighborhood

```{r}
#| warning: false
#| message: false
#| label: fig-neighborhood
#| fig_caption: true
#| fig-cap: "Bar chart shows the weighted score by the neighborhood"
#| echo: false

ggplot(analysis_data, aes(x = neighbourhood_158, y = weighted_score, fill = neighbourhood_158)) +
  geom_bar(stat = "identity") +
  labs(title = "Weighted Score by Neighbourhood", x = "Neighbourhood", y = "Weighted Score") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=5))

```

The bar chart @fig-neighborhood displays the severity of shootings and firearm discharges across various neighborhoods in Toronto, segmented by the neighbourhood_158 identifiers. Each bar represents a different neighbourhood, color-coded for visual differentiation and labeled with both the neighbourhood name and a unique identifier number.

This visualization highlights the disparity in incident severity across the city, with some neighbourhoods experiencing significantly higher weighted scores than others. Notably, certain areas show peaks which suggest hotspots of violent incidents. Such patterns are crucial for identifying regions that may require more focused law enforcement and public health interventions to mitigate the impact of firearm-related violence.

### Weighted Score by Division

```{r}
#| warning: false
#| message: false
#| label: fig-division
#| fig_caption: true
#| fig-cap: "Bar chart shows the weighted score by division"
#| echo: false

ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +
  geom_bar(stat = "identity") +
  labs(title = "Weighted Score by Division", x = "Division", y = "Weighted Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

This bar chart @fig-division illustrates the variation in the severity of shootings and firearm discharges across different police divisions in Toronto. Each division is represented by a unique color and labeled with its respective code (D11 through D55), which simplifies identification and comparison across the chart.

The visualization provides a clear depiction of how incident severity is distributed across the police divisions, with some divisions showing considerably higher weighted scores than others.

# Model Details {#sec-model-details}

## Model Description

The analysis employed a Gradient Boosting Machine (GBM) model[@gbm] using the `gbm` package in R, designed to predict the weighted scores of shootings and firearm discharges in Toronto based on multiple time-related and geographical predictors and using the `caret` package for advanced modeling techniques [@caret].

The weighted score is a computed metric that assigns double weight to deaths compared to injuries, highlighting the gravity of incidents resulting in fatalities.

## Model Setup

### Response and Predictor variables

The response variable, weighted_score, combines the effects of deaths and injuries, assigning greater importance to the former. Predictors included:

-   `occ_date`: Numeric representation of the incident date.

-   `occ_dow`: Factor variable indicating the day of the week.

-   `occ_doy`: Numeric day of the year to capture seasonal effects.

-   `occ_time_range`: Categorical variable divided into morning, afternoon, evening, and night.

-   `neighbourhood_158`: Factor variable representing different neighborhoods.

-   `division`: Factor variable representing police divisions.

### Model Configuration

The model was configured with the following parameters to control the complexity and fit of the model:

• Number of Trees: 500, providing sufficient model complexity and accuracy.

• Interaction Depth: 4, allowing interactions among up to four predictors.

• Shrinkage: 0.01, controlling the learning rate to avoid overfitting.

• Cross-validation Folds: 5, used to validate the model internally and optimize parameter selection.

## Model Performance

### Variable Importance

The summary function highlighted the relative importance of each predictor shown in @fig-importance. Notably, the neighborhood variable (neighbourhood_158) dominated the model, indicating significant spatial variation in shooting severities across Toronto. The occ_dow and division showed moderate influence, suggesting that day of the week and police division also contribute to variations in incident severity, albeit to a lesser extent.

```{r}
#| warning: false
#| message: false
#| label: fig-importance
#| fig_caption: true
#| fig-cap: "Bar chart shows the importance of the predictor variables"
#| echo: false

# Set graphical parameters
par(cex.axis = 0.4, las = 1)  # Adjust this value to change the size of the axis labels

# Plot variable importance
summary(model, n.trees = 500, plotit = TRUE)
```

### Model Performance

The model’s accuracy was assessed using the Root Mean Square Error (RMSE), calculated at 2.173830. This metric quantifies the average magnitude of the model’s prediction errors, providing a measure of predictive accuracy.

### Result Interpretation

The results underline the critical influence of geographic location (neighbourhood_158) on the severity of shooting incidents, implying that certain areas are systematically associated with higher severity scores. Temporal factors like the day of the week (occ_dow) and specific police divisions also play roles but are less pronounced compared to spatial factors.

# Discussion

## Implications

The findings from this study highlight the importance of targeted interventions in specific neighborhoods and at specific days in a week to effectively manage and mitigate the severity of shootings in Toronto. The model’s ability to pinpoint high-risk areas and times can significantly aid in deploying resources more efficiently, thereby enhancing the effectiveness of public safety measures.

## Limitations and Future Research

While the insights provided by the Gradient Boosting Machine model are invaluable, they come with inherent limitations associated with observational studies. One significant concern is the potential presence of unobserved confounders that could affect the interpretations made from the model. Variables such as unrecorded socio-economic factors, the presence of non-reported incidents, or changes in law enforcement practices over time could skew the results.

Further, the model’s dependency on historical data may not fully capture future dynamics or the impact of recent interventions. Therefore, continuous updating and validation of the model with new data are crucial for maintaining its relevance and accuracy.

Future research should aim to integrate more dynamic data sources, such as real-time crime reporting and social media analytics, which may provide more immediate indicators of changes in pattern. Additionally, exploring alternative modeling techniques, such as machine learning algorithms that can adapt over time to changes in patterns, would enhance the robustness of the findings.

## Policy Recommendations

Based on the predictive insights of the Gradient Boosting Machine model, specific tactical recommendations can be made to enhance public safety effectively:

1\. Patrol Unit Deployment:

• Increase the frequency of police patrols during late-night hours, especially around midnight to 3 AM, which the model identifies as peak times for shootings. These patrols should be intensified on weekends when the data shows a notable rise in incident severity.

• Deploy additional mobile units in neighborhoods identified as high-risk, such as West Hill, York University Heights and so on. These units can be equipped with quick-response capabilities and should be active primarily during identified peak times.

2\. Community Safety Measures:

• Establish temporary community watch programs in the most affected neighborhoods during identified peak periods. These programs could involve local volunteers working in conjunction with law enforcement to monitor and report suspicious activities.

• Implement lighting improvements and install surveillance cameras in dark alleys and poorly lit streets where incidents are frequent, as per the model’s spatial data analysis. This can help deter potential offenders and make the areas safer for residents at night.

3\. Strategic Safety Initiatives:

• Organize safety workshops and emergency response training for residents of neighborhoods with high incident rates. These workshops can focus on measures to enhance personal and collective safety, such as conflict resolution, emergency first aid, and effective communication with law enforcement.

• Partner with local businesses and community centers to fund and support extended hours of operation, providing safe spaces for youths during late hours. These centers can host activities aligned with community interests, reducing the likelihood of involvement in violence.

# Conclusion

This study offers a detailed examination of the spatial and temporal factors influencing the severity of shootings and firearm discharges in Toronto, employing a robust Gradient Boosting Machine model.

The findings highlight the critical role of specific neighborhoods and time periods in the incidence of these violent events, underscoring the potential for targeted interventions. By focusing law enforcement efforts and community resources on high-risk areas and times identified through the analysis, it is possible to significantly mitigate the impact of such incidents on public safety.

The actionable insights provided by this study not only serve as a guide for strategic policy formulation and resource allocation but also support the ongoing efforts to enhance the efficacy of public safety measures in urban environments. This approach ensures that policy responses are data-driven and tailored to the unique dynamics of the city, ultimately contributing to a safer community for all residents.

# Appendix:

## Survey and Sampling Methodology

### Overview

This study utilizes data sourced from Open Data Toronto(@OpenDataToronto), focusing on shootings and firearm discharges recorded within city limits from 2004 to 2024. Given the observational nature of the dataset, this appendix explores the methodology behind data collection, focusing on how incidents are reported and recorded, ensuring reliability and comprehensiveness of the data utilized for this analysis.

### Data Collection Method

The data collection method hinges on reports from multiple sources including police reports, hospital emergency data, and public reports. This multi-source approach helps in cross-validating the data to minimize reporting biases and errors inherent in single-source data collection. The Toronto Police Service ensures that each reported incident is followed up for additional details and context, which enhances the data’s accuracy.

### Sampling Method

While the nature of the data is not based on sampling, the comprehensiveness of data collection over a long period allows for an observational study of trends over time and across different neighborhoods. This approach mimics a census of all shooting incidents within the specified timeframe and geographical bounds, providing a detailed picture of the public safety landscape across the city.

## Addressing Biases and Limitations

### Observational Bias

Observational bias is addressed through the systematic recording of data points across all reported incidents, supplemented by stringent data verification protocols employed by the police. Data entries are standardized based on specific criteria such as incident severity, location, time, and involved parties, which are uniformly applied across all cases.

### Limitations of Observational Data

While observational data provides extensive coverage, it inherently lacks the controlled setup of experimental designs, which can introduce external variability and confounding factors. This limitation is partly mitigated through advanced statistical methods such as multivariate regression models that account for potential confounders and interaction effects among variables.

## Advanced Analytical Techniques

### Model Specification

To understand the impact of time and location on shooting severity, a Gradient Boosting Machine (GBM) model(@gbm) is employed. This section details the model’s specifications, including the handling of categorical and continuous variables, interaction terms, and the rationale behind choosing GBM over other potential models.

### Sensitivity Analysis

A sensitivity analysis is conducted to test the robustness of the model’s findings against changes in model assumptions and configurations. This analysis helps in identifying the stability of the observed effects under various scenarios and model specifications.

### Simulation Studies

Simulation studies complement the observational data analysis by testing theoretical scenarios and their probable outcomes. These simulations help in understanding potential trends in shooting incidents under different policy implementations or changes in socio-economic factors across neighborhoods.

\newpage

# References
